Towards a reproducible framework for undergraduate data science research

### 2019-12-02 Rmd/abstract.Rmd
fixed headers
updates

## Abstract

To explore social media data science as a pedagogical tool, we developed a teaching and mentoring framework to guide undergraduate researchers through individualized research projects. We demonstrated our approach by guiding two undergraduate researchers while they completed research projects. Our framework involves research question formulation, data acquisition, data analysis and visualization, and presentation of results. Our two student projects, which serve as case studies for our framework, completed all components of the individualized research projects. We found that research skills, self-confidence in research ability, and professional interest in data science increased for both students. Below, we describe our successes, lessons learned, and ideas for others to build similar frameworks for future honors projects.

>>>>

# Abstract

We develop
a mentoring framework to guide undergraduate researchers
through individualized research projects in social media data science. 
Our framework involves research question formulation, data acquisition, data 
analysis and visualization, and presentation and communication of results. Our two honors students, 
whose projects serve as case studies for our framework, completed all components of the 
individualized research projects. We found that data science research skills, self-confidence in 
research ability, and professional interest in data science increased for both 
students. We describe our successes, lessons learned, and ideas for others to 
build similar frameworks.

### 2019-12-02 Rmd/methods.Rmd
updates

Our mentored research framework begins with brainstorming scientific research ideas 
based on the student's interests. This enables us to craft a project that excites the 
student. With the results of brainstorming sessions, we (mentor and student together) 
formulate the most promising ideas into scientific hypotheses.

For the most appealing 
hypotheses, we encourage the student to translate the scientific question into a 
statistical question that may be addressed with data.

Examples may help to demonstrate our approach to identifying a statistical research 
question. One of our students had interests in acquiring and using social media posts. 
We helped her in brainstorming ideas for research involving social media like Facebook 
and Twitter. Through this brainstorming, we recognized that she had a parallel interest 
in financial markets.

study relationships between social media trends and financial market trends. We helped 
her to brainstorm related research ideas. We then

>>>>

Our mentored research framework begins with brainstorming scientific research ideas 
based on the student's interests. This enables us to craft a project that excites the 
student. With the results of brainstorming sessions, we (mentors and student together) 
formulate the most promising ideas into scientific hypotheses.

For the most appealing scientific 
hypotheses, we encourage the student to translate the scientific question into a 
statistical question that may be addressed with data. This is a crucial step in 
data science research question formulation. Skill in translating in both directions between 
scientific and statistical questions is a key communication skill that data science 
researchers offer.

Examples may help to demonstrate our approach to identifying a statistical research 
question. One of our students had interests in acquiring and using social media posts. 
We helped her in brainstorming ideas for research involving social media sources like Facebook 
and Twitter. Through this brainstorming, we recognized that she had a parallel interest 
in financial markets. Our student hypothesized that sentiment analysis of finance-related tweets 
might reflect trends in financial market index prices. On days when the market index
prices increase, sentiment analysis of finance-related tweets might reveal more use of 
positive words, while days with decreasing prices might have more negative words 
in finance-related tweets.

A second student wanted to study tweets over time and entertainment events that 
garner lots of attention in social media. We encouraged this student to develop a 
strategy for event detection from tweets over time. The rationale is that 
a big entertainment event, such as the National Football League's Super Bowl game, 
might generate enough tweets that Super Bowl-related words would appear with high 
weights in results from latent Dirichlet allocation modeling of collections of tweets 
at distinct time points. We reasoned that Super Bowl-related topics might appear during 
the Super Bowl and vanish soon after the game's conclusion.

The brainstorming sessions thus led to one set of questions for each student.
